---
title: "Lighthouse Field experiment"
output: html_document
date: "December 2023"
---

We ran a field experiment on Lady Elliot Island in 2023/2024 to test predictions about two illusions.

1. Phantom Lighthouse. We predicted that the phantom lighthouse illusion emerges at a critical distance between the observer and the source of the beam, and that this critical distance is roughly 75 m.
2. Bent Beam (height). We predicted that the Lady Elliot lighthouse beam is perceived as pointing towards the ground.

First load some packages we'll need.

```{r setup, message = FALSE}
library(here)
library(tidyverse)
library(scales)
library(brms)
library(tidybayes)
library(lme4)

knitr::opts_chunk$set(echo = TRUE)

mytheme <-  theme_classic(base_size = 10)  +  
            theme(strip.background = element_blank()) 

theme_set(mytheme)
```

```{r load_data}
# Impute with -10, 110 for people who didn't flip
thresholds <- c(-10, -10, -10, -10, 30, -10, 110, 90)
thresholds <- c(90, 90, 90, 90, 103.5, 90, 154, 146)

# number who reported beam pointing down
n_down <- 8
n_total <- length(thresholds)

d_sc <-  tibble(id = 1:n_total, data = thresholds) %>% 
  mutate(impute = if_else(data == 90 | data == 154, "imputed", "direct")) %>% 
  mutate(censored= if_else(data == 90, "left", "none")) %>% 
  mutate(censored= if_else(data == 154, "right", censored))  

d_bb <-  tibble(id = 1:n_total, data = 1)
```

We have `r n_total` participants in total. 

###  Summaries across all participants

Plot data for the phantom lighthouse condition (Figure 4b)

```{r summary_phantom}

all_results_phantom <- d_sc %>% 
  ggplot(aes(data, group=impute, color=impute, fill = impute)) +
  geom_bar() +
  scale_fill_grey(guide = "none") +
  scale_color_grey(guide = "none") +
  theme_classic()  +
  scale_x_continuous(name ="Distance (m)", 
                    breaks=  c(90, 94, 104, 116, 131, 146, 154), 
                    labels = c("<90", "94", "104", "116", "131", "146", ">154")) +
  mytheme

all_results_phantom

ggsave(here("output","figures", "thresholds_field.pdf"), plot = all_results_phantom, width = 4,height = 2)

```


### Preregistered Bayesian analyses

Because our design departed from the staircase method described in the preregistration, it did not make sense to compute a posterior distribution over the critical separation for the phantom illusion.  We therefore focus on the bent beam illusion.

```{r brms_bb, cache=TRUE}
nchain <- 4 

fit_bb  <- brm( data ~ 1, data = d_bb, family = "bernoulli", chains = nchain, iter=5000, seed = 0)
```


```{r brms_bentbeam}
bb_posterior <- fit_bb %>%
  spread_draws(b_Intercept) %>% 
  rename(intercept = b_Intercept) 

bb_posterior %>% 
  ggplot(aes(x=intercept, y = "sketch")) +
  stat_halfeye() +
  geom_vline(xintercept= 0) +
  ggtitle('Bent Beam')

fit_bb_printed <- fixef(fit_bb) %>% as_tibble(rownames="NA")
```  

The 95% credibility interval on the intercept is [`r round(fit_bb_printed$Q2.5, 2)`, `r round(fit_bb_printed$Q97.5, 2)`].

### Preregistered Frequentist analysis

For the bent beam illusion:

```{r freq_bb_binomial_test}
binom.test(8, 8, p = 0.5,
           alternative = c("greater"),
           conf.level = 0.95)
```
 